---
title: "GSDS for functional enrichment analyses"
author: "Dimitri Meistermann"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{useGSDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(7,7),
  fig.align = "center",
  cache = TRUE,
  warning=FALSE
)
```


GSDS (Gene Set Differential Scoring) is focused on functional enrichment.
The package has 3 purposes:
- provide functions to manage gene set databases
- implement well known methods to perform functional enrichment: over-representation analysis (ORA) and gene set enrichment analysis (GSEA)
- provide a new method to perform functional enrichment based on differential scoring (GSDS)

The GSDS method computes an activation score for each gene set in each sample using a PCA (similar to Pathway Level analysis of Gene Expression or PLAGE). This gene score is used  to identify gene sets that are differentially activated between two conditions.


## Installation

In a R console: 

```
install.packages("devtools")
devtools::install_github("https://github.com/DimitriMeistermann/GSDS", build_vignettes = FALSE)
```
For a manual installation of dependencies:

```
install.packages("BiocManager")
BiocManager::install( c("SummarizedExperiment","grid",
"AnnotationDbi","biomaRt","BiocParallel","ComplexHeatmap",
"circlize","coop","data.table","fgsea","ggplot2","GO.db",
"irlba","KEGGREST","matrixStats","RSQLite","ggrepel","labeling",
"scales","sparseMatrixStats","RSQLite","S4Vectors","knitr",
"rmarkdown","reactome.db","qualpalr","Rcpp","RcppArmadillo"))
```

The package is ready to load.
  
```{r setup}
library(GSDS)
```


## Over representation analysis (ORA)

For any functional enrichment method, we have to first download several gene set databases by using `getDBterms`. We will use here *kegg* and the *Gene Ontology* (biological process).
Note that you can use your own database of gene sets.

```{r getDBterms}
#species specific package will be load/asked for installation
geneSetDB<-getDBsets(species = "Human", database = c("kegg","goBP"))
geneSetDB$kegg[1:2]

```

Then we can make a simple over representation analysis.

```{r ORA}
data("DEgenesPrime_Naive")
vectorIsDE<-DEgenesPrime_Naive$isDE!="NONE";names(vectorIsDE)<-rownames(DEgenesPrime_Naive)
resEnrich<-enrich.ora(vectorIsDE,DBsets = geneSetDB)
head(resEnrich[order(resEnrich$pval,decreasing = FALSE),])

```

# functional class scoring (FCS)

The Functional class scoring family of functional enrichment is here implemented by a wrapper of `fGSEA`.
For any FCS method, we have to build first a score at the gene level.
I propose here a score for Differentially Expressed genes based on inverse of p-value with the sign of the Log2(Fold-change): $(1-p) \times \frac{Log_2FC}{|Log_2FC|} $

```{r FCS}
fcsScore<-fcsScoreDEgenes(rownames(DEgenesPrime_Naive),DEgenesPrime_Naive$pvalue,DEgenesPrime_Naive$log2FoldChange)
resEnrich<-enrich.fcs(fcsScore,DBsets = geneSetDB,returnGenes = TRUE) #return genes of the gene set in the dataframe
head(resEnrich[order(resEnrich$pval,decreasing = FALSE),])

# The result can be visualized as a volcano plot
volcanoPlot(resEnrich,effectSizeCol = "NES", adjPvalCol = "padj",labelCol = resEnrich$term)
```

We can export enrichment with the genes of each gene set in a tsv file.

```
exportEnrich(resEnrich,"test/resEnrich.tsv")
```

## Gene Set Differential Activation (GSDS) 

We can also perform an enrichment based on gene set differential activation with *GSDS*.

GSDS was developed independently from [GSVA](https://bioconductor.org/packages/release/bioc/html/GSVA.html) but is has the same idea of working on a matrix of gene set scores.
As *GSVA*, the first step is to translate the expression matrix in an activation score matrix where each column is a gene set and each row is a sample.

The particularity of *GSDS* is to perform this step by taking the first Principal Componant of the matrix $genesOfGeneSet \times samples$ from a Principal Component Analysis (PCA).
One PCA is done per gene set and the obtained values are the activation score for this particular gene set. Hence, one PCA is performed for each gene set.

The activation score matrix is then used to perform differential analysis with a linear model between two conditions.

`computeActivationScore` and `GSDS` return a `SummarizedExperiment` object, wich contain the activation score as an assay, and the metadata of the pathway, including the contributions in `rowData(object)`.

```{r GSDS}
data("bulkLogCounts")
data("sampleAnnot")
geneSetActivScore<-computeActivationScore(bulkLogCounts,DBsets = geneSetDB)
resGSDS<-GSDS(geneSetActivScore = geneSetActivScore,colData = sampleAnnot,contrast = c("culture_media","T2iLGO","KSR+FGF2"),DBsets = geneSetDB)
resGSDS.sel <- resGSDS[abs(rowData(resGSDS)$log2FoldChange) > 8 & rowData(resGSDS)$padj < 0.001,]
nrow(resGSDS.sel)
head(rowData(resGSDS.sel))
```

We can visualize the result as a heatmap with the contribution of each gene to the score.

```{r heatmapGSDS, out.width = "100%", fig.dim = c(14,10)}

GSDS.Heatmap(resGSDS.sel, colData = sampleAnnot["culture_media"])

```


